---
title: "Rat Pup Dataset"
author: "Rahul Birru"
date: "2024-03-21"
output:
  pdf_document:
    latex_engine: xelatex
  html_document:

---
# Load the Rat Pup data
```{r}
library(WWGbook)
data(ratpup)
```

**Encode the column sex of male as 0 and female as 1**
```{r}
ratpup$sex <- ifelse(ratpup$sex == "Male", 0, 1)
```

**Box plot**
```{r}
boxplot(weight ~ treatment * sex, ratpup)
## Not suited to OLS fit because of imbalance
```

# Step 1: Fit a model with a “loaded” mean structure (Model 3.1).
**Fit Model 3.1.**
```{r}
library(nlme)
model3.1.fit <- lme(weight ~ treatment + sex + litsize + treatment:sex, 
                    random = ~ 1 | litter, data = ratpup, method = "REML")
plot(model3.1.fit)
```

**Summary of model fit, along with F-tests.**
```{r}
summary(model3.1.fit)
anova(model3.1.fit)
```

**Display the random effects (EBLUPs) from the model.**
```{r}
random.effects(model3.1.fit)
```
# Step 2: Select a structure for the random effects (Model 3.1 vs. Model 3.1A).
**Fit Model 3.1A.**
```{r}
model3.la.fit <- gls(weight ~ treatment + sex + litsize + treatment:sex, data = ratpup)

anova(model3.1.fit, model3.la.fit) # Test Hypothesis 3.1.
```
# Step 3: Select a covariance structure for the residuals (Model 3.1, 3.2A, or 3.2B)
**Fit Model 3.2A.**
```{r}
model3.2a.fit <- lme(weight ~ treatment + sex + litsize + treatment:sex, 
                     random = ~1 | litter, data = ratpup, method = "REML",
                     weights = varIdent(form = ~1 | treatment))

summary(model3.2a.fit)
```

**Test Hypothesis 3.2.**
```{r}
anova(model3.1.fit, model3.2a.fit)
```
the null hypothesis does not specify that a covariance parameter lies on the
boundary of the parameter space, as in Hypothesis 3.1. Because the result of this likelihood ratio test is significant (p < .001), we choose the heterogeneous variances model (Model
3.2A) as our preferred model at this stage of the analysis.
We next test Hypothesis 3.3 to decide if we can pool the residual variances for the high and low treatment groups. To do this, we first create a pooled treatment group variable,TRTGRP

**Model 3.2B.**
```{r}
ratpup$trtgrp[ratpup$treatment == "Control"] <- 1
ratpup$trtgrp[ratpup$treatment == "Low" |ratpup$treatment == "High"] <- 2

model3.2b.fit <- lme(weight ~ treatment + sex + litsize + treatment:sex, 
                     random = ~1 | litter, data = ratpup, method = "REML",
                     weights = varIdent(form = ~1 | trtgrp))

```

**Test Hypothesis 3.3.**
```{r}
anova(model3.2a.fit, model3.2b.fit)
```
The null distribution of the test statistic in this case is a χ
2 with one degree of freedom.
Because the test is not significant (p = .27), we select the nested model, Model 3.2B, as our preferred model at this stage of the analysis.
We use a likelihood ratio test for Hypothesis 3.4 to decide whether we wish to retain the grouped heterogeneous error variances in Model 3.2B or choose the homogeneous error variance model, Model 3.1.

**Test Hypothesis 3.4.**
```{r}
anova(model3.1.fit, model3.2b.fit)
summary(model3.2b.fit)
```
The result of this likelihood ratio test is significant (p < .001), so we choose the pooled heterogeneous residual variances model, Model 3.2B, as our preferred model

# Step 4: Reduce the model by removing nonsignificant fixed effects (Model 3.2B
vs. Model 3.3, and Model 3.3 vs. Model 3.3A).
**Test Hypothesis 3.5.**
```{r}
anova(model3.2b.fit)
```
Based on the nonsignificant Type I F-test (p = .73), we delete the TREATMENT × SEX1 interaction term from the model and obtain our final model, Model 3.3.
We test Hypothesis 3.6 to decide whether the fixed effects associated with treatment are equal to zero in Model 3.3, using a likelihood ratio test based on maximum likelihood(ML) estimation then fit a nested model, Model 3.3A, without the TREATMENT term, also
using ML estimation
**Fit Model 3.3 ML.**
```{r}
model3.3.ml.fit <- lme(weight ~ treatment + sex + litsize, 
                       random = ~1 | litter, data = ratpup, method = "ML",
                       weights = varIdent(form = ~1 | trtgrp))

```

**Fit Model 3.3A ML.**
```{r}
model3.3a.ml.fit <- lme(weight ~ sex + litsize, 
                        random = ~1 | litter, data = ratpup, method = "ML",
                        weights = varIdent(form = ~1 | trtgrp))

```

**Test Hypothesis 3.6.**
```{r}
anova(model3.3.ml.fit, model3.3a.ml.fit)
```
The likelihood ratio test result is significant (p < .001), so we retain the significant fixed treatment effects in the model. We keep the fixed effects associated with SEX1 and LITSIZE without testing them, to adjust for these fixed effects when assessing the treatment effects.

We now refit our final model, Model 3.3, using REML estimation to get unbiased
estimates of the variance parameters. Note that we now specify TREATMENT as the lastterm in the fixed-effects portion of the model, so the Type I F-test reported for TREATMENT will be comparable to the Type III F-test reported

**Model 3.3: Final Model.**
```{r}
model3.3.reml.fit <- lme(weight ~ sex + litsize + treatment, 
                         random = ~1 | litter, data = ratpup, method = "REML",
                         weights = varIdent(form = ~1 | trtgrp))

summary(model3.3.reml.fit)
anova(model3.3.reml.fit)
```
```
